---
title: "Assignment-Machine.Learning"
author: "Douglas Thoms"
date: "September 19, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Project Parameters

<2000 words
=< 5 figures
-out of error rate
-accuracy compared
in htlm

how you built your model
how you used cross validation
what you think the expected out of sample error is
and why you made the choices you did




### Sections

## Prediction objective

-This group studied the quality of exercise, would they be able to classify the
quality of exercise and errors using physical sensors of movement

-The paramaters of this classification exercise are that any variables in training can
be used to correctly classify the classe variable

-The classe variable represents the type of error with "A" representing a correct
dumbell curve and "B","C","D" and "E" representing type errors in technique.

##Inputs

-The data has already been partitioned in a training and testing sets.
-No codebook has been provided but the research paper has been providing summarizing
the methodology of the study and the context of the data
-therefore, some individual variables are ambigious in their meaning such as
num_window and new_window - investigations will be need to determine these variables


## Cleaning Data, Exploratory Analysis
- The training and testing set's have different numbers of columns.  On further examination
this is because the researchers calculated more variables like minimum, maximum,
average and standard deviation.  These variables are not in the testing dataset so they cannot
be used as predictors.  These variables were removed from the training set.

- Two variables were unclear,"new_window" and "num_window".  Since there is no
codebook, the study was used.  According to the study, subjects did one set of 10 repetitions
of each "class" of exercise.  The researchers would record the data using a sliding window
ranging from 0.5 secs to 2.5 secs.  The "new_window" variables that are "yes" have the calculated
variables like minimum, maximum, etc.  This suggests these observations represent the end
of the sliding windows were all the values are calculated.  Regarding "num_window", the num_window values all consistently refer to the same classe variable.  For instance, all "num_window" 11 values correspond to classe "A".  This suggests num_window either represents a set of repetitions
or individual repetitions

-Some variables shared by both testing and training were excluded as perdictors
as they had no obvious relation to classifying the repetitions.  Removing these predictors
improved the performance of the models.

-The classified variable is discrete and not continuous.  Therefore, a simple
count bar graph was used to see the distribution of types of exercise.

```{r pressure, echo=FALSE}
plot1.obj <- ggplot(data = classe.dist, aes(x = Var2, y = Freq, fill = Var2)) +
        geom_bar(stat = "identity")
plot(plot1.obj)
```


## Features


##Evaluation

-why did I choose 5 k-folds - do these provide of out sample example?


```{r cars}
summary(cars)
```

## Including Plots

You can also embed plots, for example:



Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
