---
title: "Reproducible Research - Assignment 2 - Assessing the Health and Economic Impact of Storms"
author: "Douglas Thoms"
date: "Apr 27, 2019"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis

* Across the United States, which types of events (as indicated in the EVTYPE 
variable) are most harmful with respect to population health?  

    + Health harm will be measured in fatalities/injuries in both the total and average per event to understand the total harm these type of storms cause and the danger of individual events.  
    
    + In terms of population health, collectively heat and Tornados, Cyclone, Funnel clouds are the most dangerous events (though a combination of all miscellaneous under "OTHER" are second most dangerous).  Per Event, heat again is the biggerst danger with Tornados, Cyclone, Funnel clouds being 2nd most dangerous per event. 

* Across the United States, which types of events have the greatest economic consequences?  

    + This will be answered comparing the sum of historical crop and property damage.  This will be measured in both the total and average per event to understand the total economic damage these type of storms and the average cost for individual events.  
    
    + In terms of total economic damage, floods and hurricanes, storm surge and tropical storm are most costly.  Individual flood events don't cause much damage but the event happens frequently, while hurricanes, storm surge and tropical storms are less frequent but very costly per event.
    
# Session Info

```{r eval = TRUE}
sessionInfo()
```

# Packages Installed

```{r}
library(dplyr)
library(stringr)
library(ggplot2)
library(forcats)
library(reshape2)
library(gridExtra)
library(lubridate)
```


# Data Processing

* Data source is https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2  

* Other data documentation is https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf and https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf  

* Data was analyzed and it was found many type of events were not reported before 1993.  Therefore range of dataset analysis is 1993-2011  

* Raw data inputted with blanks entries replaced as NA  

* CROPDMGEXP and PROPDMGEXP was data cleaned according to the methodology outlined at https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html, converted into a number and multiplied with CROPDMG and PROPDMG respectively  

* TOTALDMG was calculated as a sum of CROPDMP and PROPDMG  

* EVTYPE was consolidated and data cleaned into top 6 EVTYPES for health harm and economic damage respectively, with the remaining event types being lumped into a "OTHER"  

* Two summations of health harm and economics were created - total amount of health harm and economic damage, and average health harm per event and average economic damage per event  

```{r eval=TRUE, cache=TRUE}
if(!exists("raw.data")){
        raw.data <- read.csv("repdata_data_StormData.csv.bz2", 
                             na.strings = c("",NA))

raw.data$BGN_DATE <- year(as.Date(raw.data$BGN_DATE,"%m/%d/%Y"))}

head(raw.data)

#remove columns that will not be needed
intermediate.data <- raw.data %>%
        select(EVTYPE,BGN_DATE, FATALITIES,INJURIES,PROPDMG,PROPDMGEXP,
               CROPDMG,CROPDMGEXP)

#change EXP values to exponents according to value system

#H,h,K,k,M,m,B,b,+,-,?,0,1,2,3,4,5,6,7,8, and blank-character
#H,h = hundreds = 100
#K,k = kilos = thousands = 1,000
#M,m = millions = 1,000,000
#B,b = billions = 1,000,000,000
#(+) = 1
#(-) = 0
#(?) = 0
#black/empty character = 0
#numeric 0..8 = 10

intermediate.data$PROPDMGEXP <- intermediate.data$PROPDMGEXP %>% 
        str_replace_all(c("\\+" = "1", "\\-" = "0", "\\?" = "0",
                         "[1-8]" = "10", "[hH]" = "100", "[kK]" = "1e+03", 
                         "[mM]" = "1e+06", "[bB]" = "1e+09")) %>%
        as.numeric()

intermediate.data$PROPDMGEXP[is.na(intermediate.data$PROPDMGEXP)] <- 0

intermediate.data$CROPDMGEXP <- intermediate.data$CROPDMGEXP %>% 
        str_replace_all(c("\\+" = "1", "\\-" = "0", "\\?" = "0",
                          "[1-8]" = "10", "[hH]" = "100", "[kK]" = "1e+03", 
                          "[mM]" = "1e+06", "[bB]" = "1e+09")) %>%
        as.numeric()

intermediate.data$CROPDMGEXP[is.na(intermediate.data$CROPDMGEXP)] <- 0

#multiply column with EXP

intermediate.data <- intermediate.data %>%
        transform(PROPDMG = PROPDMG * PROPDMGEXP) %>%
        transform(CROPDMG = CROPDMG * CROPDMGEXP) %>%
        mutate(TOTALDMG = CROPDMG + PROPDMG) %>%
        select(-PROPDMGEXP,-CROPDMGEXP)
        

#clean up EVTYPE entries
#subset EVTYPE =  summary entries, propdmg cropdmg values are zero - REMOVE
summary <- raw.data[grep("summary.*", raw.data$EVTYPE, ignore.case = TRUE),]

#how many hurrican entries?

hurricane <- raw.data[grep("hurricane.*", raw.data$EVTYPE, ignore.case = TRUE),]



#need to set up priority in naming to consolidate names


#check how many factors
distribution.EVTYPE <- aggregate(x = raw.data, by = list(raw.data$EVTYPE), 
                                 FUN = length)

#first, standardize same name
intermediate.data <- intermediate.data %>%
        #remove typos
        filter(!grepl("summary.*|\\?|.*apache.*", EVTYPE, ignore.case = TRUE)) %>%
        #standardize names
        transform(EVTYPE = gsub(".*hurricane.*|.*storm surge.*|.*tropical.*", 
                                "HURRICANE\nSTORM SURGE\nTROPICAL STORM", 
                                EVTYPE, ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*flood.*", "FLOOD", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*TSTM.*|.*THUNDERSTORM.*", "THUNDERSTORM", 
                                EVTYPE, ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*cyclone.*|.*tornado.*|.*funnel cloud.*|torndao", 
                                "TORNADO\nCYCLONE\nFUNNEL CLOUD", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*wind.*", "WIND", EVTYPE, 
                                ignore.case = TRUE)) %>%        
        transform(EVTYPE = gsub(".*blizzard.*", "BLIZZARD", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*microburst.*", "MICROBURST", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*freez.*|.*frost.*", "FREEZE\nFROST", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*hail.*", "HAIL", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*snow.*", "SNOW", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*light.*", "LIGHTNING", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*wint.*", "WINTER WEATHER", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*rain.*|.*shower.*|.*precipitation.*", "RAIN", 
                                EVTYPE, ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*cold.*", "COLD", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*wild.*|.*fire.*", "WILD FIRE", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*flood.*", "FLOOD", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*mud.*|.*land.*", "MUD-LAND\nSLIDE", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*heat.*|.*hot.*|.*warm.*", "HEAT", EVTYPE, 
                                ignore.case = TRUE)) %>%
        transform(EVTYPE = gsub(".*surf.*|.*current.*|.*swell.*", 
                                "CURRENT\nSURF", EVTYPE, 
                                ignore.case = TRUE)) 

distribution.check <- aggregate(BGN_DATE~EVTYPE, intermediate.data, length)

distribution.check <- distribution.check[order(-distribution.check$BGN_DATE),]

exploratory <- filter(intermediate.data, EVTYPE %in% distribution.check$EVTYPE[1:20])

exploratory <- aggregate(TOTALDMG~BGN_DATE+EVTYPE, exploratory, length)

f <- ggplot(data = exploratory, aes(x = BGN_DATE, y = TOTALDMG)) +
        geom_point(aes(colour = factor(EVTYPE)), size = 2) +
        ylab("Number of Data Points") + xlab("") +
        geom_vline(xintercept=c(1993), linetype="dotted")
```

Ran an exploratory graph and noticed that many types of event were not recorded till 1993.  Therefore only data from 1993 was used in analysis

```{r cache = TRUE}
plot(f)

intermediate.data <- filter(intermediate.data,BGN_DATE >1992)

head(intermediate.data)
```
```{r}
#check how many factors left
EVTYPE.range <- unique(intermediate.data$EVTYPE)

#Across the United States, which types of events (as indicated in the 
#EVTYPE variable) are most harmful with respect to population health?

#What columns are relevant to population health? - injuries, fatalities
#aggregate fatalities and injuries accordinng to event type
health.harm.EVTYPE <- aggregate(cbind(FATALITIES,INJURIES)~EVTYPE, 
                                intermediate.data, sum)

#count number of times same events happens
health.harm.EVTYPE.freq <- aggregate(FATALITIES~EVTYPE, 
                                     intermediate.data, length)

health.harm.EVTYPE.freq <- health.harm.EVTYPE.freq %>%
        rename(EVENT.FREQUENCY = FATALITIES)
        
health.harm.EVTYPE <- inner_join(health.harm.EVTYPE, health.harm.EVTYPE.freq, 
                                 by = "EVTYPE")


#calculate number of fatalities and injuries per event
health.harm.EVTYPE <- health.harm.EVTYPE %>%
        mutate(fatalities.percent = round(FATALITIES/sum(FATALITIES),3)) %>%
        mutate(injuries.percent = round(INJURIES/sum(INJURIES),3)) %>%
        arrange(desc(FATALITIES))
 
#rename all event types that are not top 6 into other
health.harm.EVTYPE$EVTYPE <- fct_other(health.harm.EVTYPE$EVTYPE,
                                       keep = health.harm.EVTYPE$EVTYPE[1:6],
                                       other_level = "OTHER")

#aggregate fatalities and injuries per event
health.harm.EVTYPE <- aggregate(.~EVTYPE, health.harm.EVTYPE, sum)

health.harm.EVTYPE <- health.harm.EVTYPE %>%
        mutate(fatalities.per.event = round(FATALITIES/EVENT.FREQUENCY,2)) %>%
        mutate(injuries.per.event = round(INJURIES/EVENT.FREQUENCY,2))

head(health.harm.EVTYPE)

#What columns are revelant to economic consequence? - PROPDMG, PROPDMGEXP, 
#CROPDMG, CROPDMGEXP

economic.EVTYPE <- aggregate(cbind(PROPDMG,CROPDMG,TOTALDMG)~EVTYPE, 
                             intermediate.data, sum)

#count number of times same events happens
economic.EVTYPE.freq <- aggregate(PROPDMG~EVTYPE, 
                                     intermediate.data, length)

economic.EVTYPE.freq <- economic.EVTYPE.freq %>%
        rename(EVENT.FREQUENCY = PROPDMG)

economic.EVTYPE <- inner_join(economic.EVTYPE, economic.EVTYPE.freq, 
                                 by = "EVTYPE")

#calculate total damage
economic.EVTYPE <- economic.EVTYPE %>%
        mutate(percent = round(TOTALDMG/sum(TOTALDMG),3)) %>%
        arrange(desc(TOTALDMG))

#rename all event types that are not top 6 into other
economic.EVTYPE$EVTYPE <- fct_other(economic.EVTYPE$EVTYPE,
                                       keep = economic.EVTYPE$EVTYPE[1:6],
                                       other_level = "OTHER")

economic.EVTYPE <- aggregate(.~EVTYPE, economic.EVTYPE, sum)

#aggregate all types of economic damage per event
economic.EVTYPE <- economic.EVTYPE %>%
        mutate(TOTALDMG.per.event = round(TOTALDMG/EVENT.FREQUENCY,0)) %>%
        mutate(PROPDMG.per.event = round(PROPDMG/EVENT.FREQUENCY,0)) %>%
        mutate(CROPDMG.per.event = round(CROPDMG/EVENT.FREQUENCY,0))

head(economic.EVTYPE)
```

# RESULTS
```{r eval=TRUE}
#Across the United States, which types of events (as indicated in the EVTYPE 
#variable) are most harmful with respect to population health?


#create tall dataframe for charting purposes
health.harm.total <- health.harm.EVTYPE %>%
        select(EVTYPE,FATALITIES,INJURIES) %>%
        melt(id.vars = "EVTYPE", measure.vars = c("FATALITIES","INJURIES")) 

health.harm.total <- health.harm.total[order(health.harm.total$variable, 
                                             -health.harm.total$value),] 

#set descending order of event type factors
EVTYPE.ORDER <- health.harm.total %>%
        filter(variable == "FATALITIES") %>%
        select(-variable,-value)
        
health.harm.total$EVTYPE <- factor(health.harm.total$EVTYPE, 
                                   levels = EVTYPE.ORDER$EVTYPE)

health.harm.total$variable <- factor(health.harm.total$variable, 
                                     levels = c("INJURIES","FATALITIES"))

#graph total fatalities and injuries
g <- ggplot(data=health.harm.total, aes(x=EVTYPE, y=value, 
                                        fill=factor(variable))) +
        geom_bar(colour="black", stat="identity",
                 position=position_dodge(),
                 size=.3, width = .9) + 
        theme_bw() +
        scale_fill_hue(name="Harm") + 
        xlab("") + ylab("") +
        ggtitle("Total Fatalities and Injuries") +
        theme(axis.text.x = element_text(angle = 90, hjust = 0.75, vjust = 0.5, 
                                         size=9,color="darkblue"))


health.harm.per.event <- health.harm.EVTYPE %>%
                select(EVTYPE,fatalities.per.event,injuries.per.event) %>%
                melt(id.vars = "EVTYPE", measure.vars = c("fatalities.per.event",
                                                          "injuries.per.event"))

health.harm.per.event$EVTYPE <- factor(health.harm.per.event$EVTYPE, 
                                       levels = EVTYPE.ORDER$EVTYPE)
health.harm.per.event$variable <- factor(health.harm.per.event$variable, 
                                         levels = c("injuries.per.event",
                                                    "fatalities.per.event"))

#graph fatalities and injuries per event 
h <- ggplot(data=health.harm.per.event, aes(x=EVTYPE, y=value, 
                                            fill=factor(variable))) +
        
                geom_bar(colour="black", stat="identity",
                         position=position_dodge(),
                         size=.3) +                       
                scale_fill_hue(name="Harm") + xlab("") + ylab("") +
        ggtitle("Fatalities and Injuries Per Event") + theme_bw() +
        theme(axis.text.x = element_text(angle = 90, hjust = 0.75, vjust = 0.5, 
                                         size=9,color="darkblue"))

#print plot
grid.arrange(g,h, ncol=1)


        
#Across the United States, which types of events have the greatest economic consequences?

#create tall dataframe for charting purposes
economic.total <- economic.EVTYPE %>%
                select(EVTYPE,PROPDMG,CROPDMG,TOTALDMG) %>%
                melt(id.vars = "EVTYPE", measure.vars = c("PROPDMG","CROPDMG","TOTALDMG"))

economic.total <- economic.total[order(economic.total$variable, -economic.total$value),] 

#set factor order according to largest total damage
EVTYPE.ORDER <- economic.total %>%
        filter(variable == "TOTALDMG") %>%
        select(-variable,-value)

economic.total <- filter(economic.total, variable != "TOTALDMG")

economic.total$EVTYPE <- factor(economic.total$EVTYPE, levels = EVTYPE.ORDER$EVTYPE)       
economic.total$variable <- factor(economic.total$variable, levels = c("CROPDMG",
                                                                      "PROPDMG"))
#graph total prop and crop damage     
i <- ggplot(data=economic.total, aes(x=EVTYPE, y=value/1e+09, fill=factor(variable))) +
                geom_bar(colour="black", stat="identity",
                         position="stack", size=.3) +                       
                scale_fill_hue(name="Damage") + xlab("") + ylab("") +
        ggtitle("Total Crop and Property Damage (Billions)") + theme_bw() +
        theme(axis.text.x = element_text(angle = 90, hjust = 0.75, vjust = 0.5, 
                                         size=9,color="darkblue"))
    
        
economic.per.event <- economic.EVTYPE %>%
                select(EVTYPE,PROPDMG.per.event,CROPDMG.per.event) %>%
                melt(id.vars = "EVTYPE", measure.vars = c("PROPDMG.per.event",
                                                          "CROPDMG.per.event"))
        
economic.per.event <- economic.per.event[order(economic.per.event$variable, 
                                               -economic.per.event$value),] 

#set order of factors in descending damage cost
economic.per.event$EVTYPE <- factor(economic.per.event$EVTYPE, 
                                    levels = EVTYPE.ORDER$EVTYPE)       
economic.per.event$variable <- factor(economic.per.event$variable, 
                                      levels = c("CROPDMG.per.event",
                                                 "PROPDMG.per.event"))

#graph total prop and crop damage per event       
j <- ggplot(data=economic.per.event, aes(x=EVTYPE, y=value/1e+06, 
                                         fill=factor(variable))) +
        
                geom_bar(colour="black", stat="identity",
                         position="stack", size=.3) +                       
                scale_fill_hue(name="Damage") + xlab("") + ylab("") +
        ggtitle("Total Crop and Property Damage per Event (Millions)") +    
                theme_bw() +
        theme(axis.text.x = element_text(angle = 90, hjust = 0.75, vjust = 0.5, 
                                         size=9,color="darkblue"))

        
        
#print plot        
grid.arrange(i,j, ncol=1)
```
