---
title: "Reproducible Research - Assignment 1"
author: "Douglas Thoms"
date: "April 9, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.width = 8)
```

##Packages Installed

```{r echo = FALSE}
##Packages used
require('plyr')
require('dplyr')
require('zoo')
require('lubridate')
require('lattice')
```



## Loading and preprocessing the data

```{r}
rawData <- read.csv("activity.csv")
```



## What is mean total number of steps taken per day?

Steps per day was calculated using the following code:

```{r}

stepsPerDay <- aggregate(x = rawData$steps,
                           by = list(rawData$date), sum)
colnames(stepsPerDay) <- c("date","steps.per.day")

```

The historgram below shows the distribution:

```{r eval = TRUE}
hist(stepsPerDay$steps.per.day, breaks = seq(from = 0, to = 25000, by = 1250), 
     xlab = "", main = "Steps per day")

```

The median steps per day was `r median(stepsPerDay$steps.per.day, na.rm = TRUE)`.

The mean steps per day was `r as.character(round(mean(stepsPerDay$steps.per.day, na.rm = TRUE),1))`

```{r eval = TRUE}
median(stepsPerDay$steps.per.day, na.rm = TRUE)
mean(stepsPerDay$steps.per.day, na.rm = TRUE)
```



## What is the average daily activity pattern?

The following code was used  to find the average steps per interval.

```{r eval = TRUE}

meanStepsInterval <- aggregate(x = rawData$steps,
                         by = list(rawData$interval), mean, na.rm = TRUE)
colnames(meanStepsInterval ) <- c("interval","mean.steps.per.interval")

plot(y = meanStepsInterval $mean.steps, x= meanStepsInterval $interval, type = "l", 
     xaxt = "n", xlab = "5-min Interval", ylab = "Average Steps", main = "Mean.Steps")

axis(side = 1, at = c(seq(from = 0, to = 2355, by = 60)))
```

The following code was used to find the interval with the maximum steps per interval.

```{r eval = TRUE}
MaxmeanStepsInterval <- match(max(meanStepsInterval $mean.steps),
                                meanStepsInterval $mean.steps)

meanStepsInterval$interval[MaxmeanStepsInterval]
```

The interval with the highest average steps was `r meanStepsInterval$interval[MaxmeanStepsInterval]`



## Imputing missing values

```{r eval= TRUE}
length(rawData$steps[is.na(rawData$steps)])
```

There are `r length(rawData$steps[is.na(rawData$steps)])` NA intervals.

A new dataset was created by replacing the first day of NA values with the average steps per interval.
The other NAs are replaced by interpolating values from the same interval of the prior and following day using zoo::na.approx.
Here is the code:

```{r}
replacedNA <- rawData
replacedNA$steps[1:288] <- meanStepsInterval $mean.steps.per.interval 

replacedNA <- ddply(replacedNA, .(interval), transform, steps=na.approx(steps, rule=2))
replacedNA <- replacedNA[order(replacedNA$date,replacedNA$interval), ]

stepsPerDayNAremoved <- aggregate(x = replacedNA$steps,
                         by = list(replacedNA$date), sum)
colnames(stepsPerDayNAremoved) <- c("date","steps.per.day")
```

The following code was used to create the plots below:
```{r eval = TRUE}
par(mfrow= c(1,2))

hist(stepsPerDay$steps.per.day, breaks = seq(from = 0, to = 25000, by = 1250), 
     xlab = "", main = "Steps per day", ylim = c(0,14))

hist(stepsPerDayNAremoved$steps.per.day, 
     breaks = seq(from = 0, to = 25000, by = 1250), xlab = "", 
     main = "Steps per day NA removed", ylim = c(0,14))
```

The plots are very similar in distribution and the NA-replaced dataset has mainly increased values in the middle of the distribution.

The follow code was used to recalculate the median and mean with the NAs substitued with interpolated values:

```{r eval = TRUE}
median(stepsPerDay$steps.per.day, na.rm = TRUE)
round(mean(stepsPerDay$steps.per.day, na.rm = TRUE),1)
```

The median for the NA subsituted dataset is `r median(stepsPerDay$steps.per.day, na.rm = TRUE)`.
The mean for the NA subsituted dataset is `r as.character(round(mean(stepsPerDay$steps.per.day, na.rm = TRUE),1))`.  This mean and median is nearly identical to the original dataset with NA values.