---
title: "Statistical Inference - Assignment Part 2"
author: "Douglas Thoms"
date: "June 9, 2019"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.height = 3, fig.width = 8, message = FALSE)
```

### Library
```{r echo=TRUE}
library(dplyr)
library(ggplot2)
```

### Overview
This report will explore ToothGrowth Data and whether different delivery methods and doses are statistically related to differences in tooth growth rate.  Confidence Intervals and Hypothesis tests will be used to see if there is significance difference.


### Data 
"The response is the length of odontoblasts (cells responsible for tooth growth) 
in 60 guinea pigs. Each animal received one of three dose levels of vitamin C (0.5, 1, and 2 mg/day) 
by one of two delivery methods, orange juice or ascorbic acid (a form of vitamin C and coded as VC).""
(https://stat.ethz.ch/R-manual/R-devel/library/datasets/html/ToothGrowth.html)
    
### Hypothesis

Null hypothesis: The average tooth growth is no different regardless of dose or delivery method used.  
$H_o: \mu_o = \mu_a$, $H_a: \mu_o \neq \mu_a$ 

### Data Exploration

The data is structured in the following way.
```{r eval=TRUE}
df <- ToothGrowth
head(df)
```


```{r}
num.NA.obs <- length(df[complete.cases(df) == FALSE,]$len)
```
No data quality issues were detected.  There are `r length(df[complete.cases(df) == FALSE,]$len)` missing or non-number values.

```{r eval=TRUE}
#aggregate to see differences
ag.mean.df <- aggregate(.~supp+dose, df, mean)
ag.mean.df <- ag.mean.df %>%
        rename(mean.length = len)
ag.mean.df$mean <- round(ag.mean.df$mean, digits = 2)

ag.sd.df <- aggregate(.~supp+dose, df, sd)
ag.sd.df <- ag.sd.df %>%
        rename(sd.length = len)
ag.sd.df$sd <- round(ag.sd.df$sd, digits = 2)

ag.mean.sd <- inner_join(ag.mean.df,ag.sd.df)

ag.mean.sd <- ag.mean.sd[with(ag.mean.sd, order(supp, dose)),]
print(ag.mean.sd)
```
The variance of the different doses and delivery methods are not consistent as the table shows below.  This will affect our confidence interval and p-value calculation method - the Two Sample Welch t-test will be used.  

This is further confirmed by graphing the distribution of values.
```{r eval=TRUE}
df$dose <- as.factor(df$dose)
a <- ggplot(df, aes(x=dose, y=len)) + geom_boxplot() +
        facet_grid(.~supp)
plot(a)
```

### Data Manipulation

Data is subsetted into groups.
```{r}
small.dose.len <- df %>%
        filter(len, dose == 0.5) 
med.dose.len <- df %>%
        filter(len, dose == 1) 
high.dose.len <- df %>%
        filter(len, dose == 2) 

OJ.small.len <- small.dose.len %>%
        filter(len, supp == "OJ")
VC.small.len <- small.dose.len %>%
        filter(len, supp == "VC")
OJ.med.len <- med.dose.len %>%
        filter(len, supp == "OJ")
VC.med.len <- med.dose.len %>%
        filter(len, supp == "VC")
OJ.high.len <- high.dose.len %>%
        filter(len, supp == "OJ")
VC.high.len <- high.dose.len %>%
        filter(len, supp == "VC")
```

### Hypothesis Tests

A two-tailed 95% confidence interval will be used since the hypothesis are as follows    
$H_o: \mu_o = \mu_a$, $H_a: \mu_o \neq \mu_a$  
Therefore, $\alpha=0.025$  

#### Delivery Method Test
The delivery methods were compared against each other at different doses.
```{r evalu=TRUE, message = TRUE}
#OJ 0.5 vs VC 0.5
t.test(len ~ supp, data = rbind(OJ.small.len, VC.small.len))
#OJ 1 vs VC 1
t.test(len ~ supp, data = rbind(OJ.med.len, VC.med.len))
#OJ 2 vs VC 2
t.test(len ~ supp, data = rbind(OJ.high.len, VC.high.len))
```
 * At 1 mg/day or lower, the p-value was considered significant and the confidence interval did not include zero.  
 * 2 mg/day we failed to reject the null hypothese as p-value was too large and the confidence interval include 0.    
 * The null hypothesis that delivery method is statistically irrelevant can be rejected at 1 mg/day and lower, but cannot be rejected at 2 mg/day.   

#### Dose Level Test
The dose were compared against each other using the same delivery methods.
```{r evalu=TRUE, message = TRUE}
#OJ 0.5 vs OJ 0.1
t.test(len ~ dose, data = rbind(OJ.small.len, OJ.med.len))
#OJ 1 vs OJ 2
t.test(len ~ dose, data = rbind(OJ.med.len, OJ.high.len))
#OJ 0.5 vs OJ 2
t.test(len ~ dose, data = rbind(OJ.small.len, OJ.high.len))
#VC 0.5 vs VC 0.1
t.test(len ~ dose, data = rbind(VC.small.len, VC.med.len))
#VC 1 vs VC 2
t.test(len ~ dose, data = rbind(VC.med.len, VC.high.len))
#VC 0.5 vs VC 2
t.test(len ~ dose, data = rbind(VC.small.len, VC.high.len))
```
 * In all tests, the results were considered all relevant as p-values were less than $\alpha$, except for medium and high doses of Orange Juice.
 * The difference between 1 mg/day and 2 mg/day doses of Orange Juice has zero in the confidence interval so we cannot reject the null hypothesis in this case
 * In all other cases we can reject the null hypothesis that dose is statistically unassociated with differences in tooth growth.

### Conclusion

The following assumptions were made about the data:  

 * Observations are iid variables - independent and identical distributions
 * Data is not paired
 * Data groups variance are not the same

The following conclusions can be drawn from the hypothesis testing:  

 * Choice of delivery method is statistically relevant up to 1 mg/day.  There is a statistical association between teeth growth difference and delivery method of vitamin C at these doses.
 * Choice of delivery method is irrelevant for teeth growth difference at 2 mg/day.  There is no statistical association between teeth growth difference and delivery method of vitamin C at 2 mg/day.
 * Amount of dose is statistically relevant for teeth growth regardless of delivery method in most cases.  There is a statistical association between teeth growth difference and doses.
 * There is no statistical association in teeth growth association and 1 mg/day and 2 mg/day doses of Orange Juice.