---
title: "Statistical Inference - Assignments"
author: "Douglas Thoms"
date: "June 9, 2019"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Synopsis
    
# Session Info

```{r eval = TRUE}
sessionInfo()
```

# Packages Installed

```{r}
library(dplyr)
library(stringr)
library(ggplot2)
library(forcats)
library(reshape2)
library(gridExtra)
library(lubridate)
```


# Data Processing

* Data source is https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2  

* Other data documentation is https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf and https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2FNCDC%20Storm%20Events-FAQ%20Page.pdf  

* Data was analyzed and it was found many type of events were not reported before 1993.  Therefore range of dataset analysis is 1993-2011  

* Raw data inputted with blanks entries replaced as NA  

* CROPDMGEXP and PROPDMGEXP was data cleaned according to the methodology outlined at https://rstudio-pubs-static.s3.amazonaws.com/58957_37b6723ee52b455990e149edde45e5b6.html, converted into a number and multiplied with CROPDMG and PROPDMG respectively  

* TOTALDMG was calculated as a sum of CROPDMP and PROPDMG  

* EVTYPE was consolidated and data cleaned into top 6 EVTYPES for health harm and economic damage respectively, with the remaining event types being lumped into a "OTHER"  

* Two summations of health harm and economics were created - total amount of health harm and economic damage, and average health harm per event and average economic damage per event  

# Simulations
Include English explanations of the simulations you ran, with the accompanying R code. Your explanations should make clear what the R code accomplishes.

```{r}
lambda = 0.2

#calculate population parameters
pop.mean <- 1/lambda
pop.sd <- 1/lambda
pop.variance <- pop.sd^2

hist(rexp(1000))
abline(v = 5, col = "green", lwd = 3)

expo_mean <- NULL
for (i in 1 : 1000) expo_mean = c(expo_mean, mean(rexp(40, 0.2)))
hist(expo_mean)
abline(v = 5, col = "green", lwd = 3)
```

# Sample Mean versus Theoretical Mean
Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.
# Sample Variance versus Theoretical Variance
Include figures (output from R) with titles. # Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.
#Distribution
Via figures and text, explain how one can tell the distribution is approximately normal.